\Large The state learner -- a super learner for right-censored data

### 2024-12-19 manuscript/manuscript.tex
restructure only one version
edit intro, focus on other approaches

%% Line spacing
\linespread{1.65}

@@

Machine learning based on right-censored data commonly uses the
partial log-likelihood as a loss function
\citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
However, this loss function does not work well with data splitting
(cross-validation) because the partial log-likelihood loss assigns an
infinite value when applied to a test set time point which does not
occur in the learning set as soon as the learner predicts piece-wise
constant cumulative hazard functions. This is the case for prominent
survival learners including the Kaplan-Meier estimator, the random
survival forest, and the semi-parametric Cox regression model.  When a
proportional hazards model is assumed, the baseline hazard function
can be profiled out of the likelihood \citep{cox1972regression}.  The
cross-validated partial log-likelihood loss \citep{verweij1993cross}
has therefore been suggested as a loss function for super learning
which however restricts the library of learners to include only Cox
proportional hazards models \citep{golmakani2020super}.
% A lesser
% known fact is that a super learner constructed with the partial
% log-likelihood loss implicitly depends on the censoring distribution
% \citep{hjort1992inference,whitney2019comment}.

Alternative approaches for super learning with right-censored data use
an inverse probability of censoring weighted (IPCW) loss function
\citep{graf1999assessment,van2003unicv,molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked},
% An IPCW loss function is attractive because the associated risk does
% not depend on the censoring distribution but describes a feature of
% the population of interested. Similar results can be obtained using
censoring unbiased transformations
\citep{fan1996local,steingrimsson2019censoring}, or pseudo-values
\citep{andersen2003generalised,mogensen2013random,sachs2019ensemble}.
All these methods rely on an estimator of the censoring distribution,
and their drawback is that this estimator has to be pre-specified.
% When the data-generating mechanism is complex and not
% well-understood, pre-specification of the censoring distribution is a
% challenge.
% Inverse probability of censoring weighted loss has been studied for
% decades \citep{graf1999assessment, hothorn2006survival,
  % gerds2006consistent}. A disadvantage of inverse probability of
% censoring weighted loss estimators is that they require a
% pre-specified model for the censoring distribution.
% Recently \cite{westling2021inference} proposed a procedure which
% iterates between super learning of the censoring distribution and
% super learning of the event time distribution.
An approach which avoids a pre-specified censoring model was proposed
independently by \cite{han2021inverse} and
\cite{westling2021inference}. In both articles, the authors suggest to
iterate between learning of the outcome model and learning of the
censoring model using IPCW loss functions. However, no general
theoretical guarantees seem to exist for this procedure, and it has
not yet been extended to the situation with competing risks.

The state learner algorithm can output a medical risk prediction model
\citep{gerds2021medical} which predicts the probability of an event based on
covariates in the presence of competing risks. The other application is in
targeted learning where conditional event probabilities occur as
high-dimensional nuisance parameters which need to be estimated at a certain
rate \citep{van2011targeted, rytgaard2021estimation, rytgaard2022targeted}. For
the asymptotic bias term of targeted estimator, which uses the state learner to
estimate nuisance parameters, we show that a second order product structure
holds. We illustrate both applications of the state learner with prostate cancer
data \citep{kattan2000pretreatment}.

@@

\begin{figure}
  \centering%
  \includeFigCond[width=.9\linewidth]{real-data-target.pdf}
  \caption[]{Estimates of the average treatment effect of hormone therapy on the
    risk of tumor recurrence and death obtained from the prostate cancer study
    analysed by \cite{kattan2000pretreatment}. The estimates are based on the
    estimator defined in equation~(\ref{eq:one-step-comp-ate}) with pointwise
    95\% confidence intervals calculated as described in
    Section~\ref{sec:targeted-learning}. The estimates of the nuisance
    parameters are provided by the state learner.}
  \label{fig:zelefski-real-target}
\end{figure}

>>>>

%% Line spacing
% \linespread{1.65}

@@

The state learner algorithm can output a medical risk prediction model
\citep{gerds2021medical} which predicts the probability of an event based on
covariates in the presence of competing risks. Another important application,
which motivated our contribution, is in targeted learning where conditional
event probabilities occur as high-dimensional nuisance parameters which need to
be estimated at a certain rate \citep{van2011targeted, rytgaard2021estimation,
  rytgaard2022targeted}. The state learner is particularly well-suited for this
application as it simultaneously provides estimates for both the outcome and the
censoring, both of which are needed for targeted learning. For the asymptotic
bias term of a targeted estimator, which uses the state learner to estimate
nuisance parameters, we show that a second order product structure holds. We
illustrate both applications of the state learner with prostate cancer data
\citep{kattan2000pretreatment}.

There are many existing super learners for right-censored data. We briefly
review them here and contrast our suggestion. Machine learning based on
right-censored data commonly uses the partial log-likelihood as a loss function
\citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
This loss function is also suggested for super learning with right-censored data
by \cite{polley2011-sl-cens}, where data is assumed to be observed in discrete
time. However, the partial log-likelihood loss does not work well with data
splitting (cross-validation) in continuous time because this loss assigns an
infinite value when applied to a test set time point which does not occur in the
learning set as soon as the learner predicts piece-wise constant cumulative
hazard functions. This is the case for prominent survival learners including the
Kaplan-Meier estimator, the random survival forest, and the semi-parametric Cox
regression model, which hence cannot be included in the library of the super
learner proposed by \cite{polley2011-sl-cens}.

When a proportional hazards model is assumed, the baseline hazard function can
be profiled out of the likelihood \citep{cox1972regression}. The cross-validated
partial log-likelihood loss \citep{verweij1993cross} has therefore been
suggested as a loss function for super learning \citep{golmakani2020super}. This
choice of loss function restricts the library of learners to include only Cox
proportional hazards models, and hence excludes many learners such as, e.g.,
random survival models, additive hazards models and accelerated failure time
models.
% A lesser
% known fact is that a super learner constructed with the partial
% log-likelihood loss implicitly depends on the censoring distribution
% \citep{hjort1992inference,whitney2019comment}.

Other attempts at super learning with right-censored data use an inverse
probability of censoring weighted (IPCW) loss function
\citep{graf1999assessment,van2003unicv,molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked},
% An IPCW loss function is attractive because the associated risk does
% not depend on the censoring distribution but describes a feature of
% the population of interested. Similar results can be obtained using
censoring unbiased transformations
\citep{fan1996local,steingrimsson2019censoring}, or pseudo-values
\citep{andersen2003generalised,mogensen2013random,sachs2019ensemble}.
All these methods rely on an estimator of the censoring distribution,
and their drawback is that this estimator has to be pre-specified.
% When the data-generating mechanism is complex and not
% well-understood, pre-specification of the censoring distribution is a
% challenge.
% Inverse probability of censoring weighted loss has been studied for
% decades \citep{graf1999assessment, hothorn2006survival,
  % gerds2006consistent}. A disadvantage of inverse probability of
% censoring weighted loss estimators is that they require a
% pre-specified model for the censoring distribution.
% Recently \cite{westling2021inference} proposed a procedure which
% iterates between super learning of the censoring distribution and
% super learning of the event time distribution.

A recent approach which avoids a pre-specified censoring model was proposed
independently by \cite{han2021inverse} and \cite{westling2021inference}. In both
articles, the authors suggest to iterate between learning of the outcome model
and learning of the censoring model using IPCW loss functions. However, no
general theoretical guarantees seem to exist for this procedure. For instance,
the procedure is not guaranteed to converge to the true data-generating
mechanism \citep{munch2024thesis}.

Contrary to the super learners considered by \cite{polley2011-sl-cens} and
\cite{golmakani2020super}, the state learner poses no restrictions on the type
of learners that can be included in the library. Contrary to IPCW, censoring
unbiased transformations, or pseudo-value based super learners, the state
learner does not require a pre-specified censoring distribution, but constructs
a super learner for both the outcome and the censoring distribution. Contrary to
the proposals by \cite{han2021inverse} and \cite{westling2021inference}, the
state learner is backed by theoretical guarantees.

@@

\begin{figure}
  \centering%
  \includeFigCond[width=1\linewidth]{real-data-target.pdf}
  \caption[]{Estimates of the average treatment effect of hormone therapy on the
    risk of tumor recurrence and death obtained from the prostate cancer study
    analysed by \cite{kattan2000pretreatment}. The estimates are based on the
    estimator defined in equation~(\ref{eq:one-step-comp-ate}) with pointwise
    95\% confidence intervals calculated as described in
    Section~\ref{sec:targeted-learning}. The estimates of the nuisance
    parameters are provided by the state learner.}
  \label{fig:zelefski-real-target}
\end{figure}
